---
title: "a03 cleaning data"
author: Tokay Alberts
editor: visual
on:
  workflow_dispatch:
date: March 12, 2026
---

```{r}
#|warning: false
library(dplyr)
library(tidyr)
library(readr)
```

## Introduction

The salmon fisheries industry is a massive, multi-billion dollar industry that deeply affects many aspects of peoples lives and ecosystems. With climate change, overfishing is becoming more of a concern as populations struggle to rebound from threats at several levels. Long-term occurrence data is thus essential to monitoring trends in population take and assessing sustainability. Salmon are also considered a keystone species, which further support other species and can serve as indicators of ecosystem health.

## Data

The data in this assignment was downloaded from the KNB repository and includes Alaska commercial salmon catch by region from 1886-1997. It was downloaded at this link on March 12th, 2026: <https://knb.ecoinformatics.org/view/df35b.304.2>

This data was generated from a fish ticket system, which reports all salmon that are sold.

```{r}
catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"))

```

## Analysis

```{r}
#data exploration
## First 6 lines of the data frame
head(catch_original)
## Summary of each column of data
summary(catch_original)
```

```{r}
#filter for the columns you want
catch_data <- catch_original %>%
    select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_data)
```

```{r}
dplyr::glimpse(catch_data)
# changing chinook from character to numeric
catch_clean <- catch_data %>%
    mutate(Chinook = as.numeric(Chinook))
head(catch_clean)
```

```{r}
i <- which(is.na(catch_clean$Chinook))
i
#view row 401
catch_data[i,]

#change the i to a 1
catch_clean <- catch_data %>%
    mutate(Chinook = if_else(condition = Chinook == "I", 
                             true = "1", 
                             false = Chinook),
           Chinook = as.numeric(Chinook))

##check
catch_clean[i, ]
```

```{r}
catch_long <- catch_clean %>% 
    #pivot longer all columns except Region and Year
    pivot_longer(
        cols = -c(Region, Year),
        names_to = "species",
        values_to = "catch"
    )

head(catch_long)

# change catch to thousands
catch_long <- catch_long %>%
    rename(catch_thousands = catch)

catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000)

# remove catch thousands column
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000) %>%
    select(-catch_thousands)

head(catch_long)
```

```{r}
#summarizing means by region
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch))

head(mean_region)
```

```{r}
#filtering
sse_catch <- catch_long %>%
    filter(Region == "SSE")

head(sse_catch)
# sorting
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(mean_catch)

head(mean_region)
#splitting by location
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAU-039"))

sites_df %>%
    separate(site, c("island", "site_number"), "-")
```

## Discussion

This analysis demonstrates the importance of data cleaning and filtering to manipulate variables for better readability. Replacing incorrect values using functions is essential for reproducibility and helps avoid future headaches. This data also lays out the most productive regions of salmon catch across several decades.
